---
title: "Storing Mass Spectrometry Data in MongoDB Databases"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Storing Mass Spectrometry Data in MongoDB Databases}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{MsBackendMongoDb}
    %\VignetteDepends{MsBackendMongoDb,BiocStyle,msdata,mzR,mongolite}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

# Introduction

The `r Biocpkg("Spectra")` Bioconductor package provides a flexible and
expandable infrastructure for handling and processing mass spectrometry (MS)
data in R. The package supports interchangeable use of different *backends* that
provide additional file format support or different ways to store and
*represent* MS data. The *MsBackendMongoDb* package expands *Spectra* enabling
to store and retrieve MS data from [MongoDB](https://www.mongodb.com/)
databases. MongoDB is a *NoSQL* database for storing structured, semi-structured
and unstructured data.

```{r, echo = FALSE}
#' Only evaluate code if a running mongodb server is available on localhost
HAVE_MONGO <- TRUE
```

# Installation and system requirements

*MsBackendMongoDb* along with all required dependencies can be installed with:

```{r, eval = FALSE}
install.packages("BiocManager")
install.packages("remotes")
BiocManager::install("RforMassSpectrometry/MsBackendMongoDb")
```

For using *MsBackendMongoDb*, access to a MongoDB, either running locally or
remotely, is required. See the official [MongoDB](https://www.mongodb.com/) web
page for information on installation and setup of a MongoDB database server.

# Storing MS data in a MongoDB database

In this section we store MS data from two mzML files into an empty MongoDB
database. We create below a `Spectra` object representing the MS data from these
two data files.

```{r}
library(Spectra)
library(msdata)
fls <- dir(system.file("sciex", package = "msdata"), full.names = TRUE)

sp <- Spectra(fls)
sp
```

This `Spectra` object has thus `r length(sp)` spectra. We next change the
MS data *storage* mode for this object to a MongoDB database. We therefore
establish below a connection to a local MongoDB server. The
`connectMsBackendMongoDb()` function connects to a database with the name
specified with parameter `db` and creates the required MongoDB *collections* to
store the MS data. The database will be created automatically if it does not
already exist. By setting parameter `clean = TRUE` we ensure that any previous
database with the same name as defined with `db` gets first erased. Thus, this
parameter should be used with caution!

```{r}
#' connect to a MongoDB database
library(MsBackendMongoDb)
mcon <- connectMsBackendMongoDb(
    db = "ms_backend_mongodb",
    url = "mongodb://localhost:27017",
    clean = TRUE)
```

We can next use the `setBackend()` function to transfer the data from the
original mzML files to the MongoDB database.

```{r}
sp_mng <- setBackend(sp, backend = MsBackendMongoDb(), dbcon = mcon)
```

The memory requirements for a `Spectra` object with a `MsBackendMongoDb` are
very low since only the IDs of the spectra are kept in the main memory while all
data is stored in the database:

```{r}
print(object.size(sp_mng), units = "MB")
```

Due to the design of the `Spectra` class, data can be retrieved irrespectively
of where or how the data is stored. For `Spectra` objects using the
`MsBackendMongoDb` backend, the data is extracted *on-the-fly* from the MongoDB
database. Below we get the retention time values from the object and show the
first 6 values.

```{r}
rtime(sp_mng) |>
    head()
```

Similarly, we can extract the *m/z* values of all spectra from the data
set. Also in this case, the data is extracted from the database.

```{r}
mz(sp_mng)
```

Subset operations will only subset the spectra IDs in the backend, but not
affect the data stored in the MongoDB. Below we subset the `Spectra` object to
spectra with index 13, 5, 20 and 21:

```{r}
sp_mng <- sp_mng[c(13, 5, 20, 21)]
sp_mng
```

Any operation on this subset `Spectra` object will only retrieve data from the 4
selected spectra.

```{r}
rtime(sp_mng)
```

As mentioned above, the MongoDB database still contains the full data. We can
thus use the `reset()` function to *restore* the `Spectra` object to its
original state:

```{r}
sp_mng <- reset(sp_mng)
sp_mng
```

Finally, the data available from the `Spectra` object using the
`MsBackendMongoDb` is identical to the original `Spectra` object with the data
from the original mzML files. We compare below the retention times reported by
both `Spectra` objects.

```{r}
all.equal(rtime(sp), rtime(sp_mng))
```

Also the *m/z* values are identical - with the exception of the names of the
returned `list`.

```{r}
all.equal(mz(sp), unname(mz(sp_mng)))
```

# Session information

```{r}
sessionInfo()
```
